Hybrid Threads for the Parrot Virtual Machine

### 2012-02-24 Paper.tex
Document the functions in src/scheduler.c

\section{Runloops}

@@

\chapter{Implementation of Hybrid Threads (15)}

@@

Runloops are created whenever a part of Parrot C internals start to execute bytecode. This could be simply at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback. An exception handler could also be such a callback. Each runloop has a unique id. These ids are numbered from 0 and simply increased by 1 whenever a new runloop is created.

@@

To mitigate this, the \textit{reset\_runloop\_id\_counter(interp)} function is used to reset the global runloop id counter back to 0 when resuming a task. This way the task retains the same runloop id over its life time.

If Parrot were to preempt a task while it is executing a nested runloop, it would have to somehow capture not only the interpreter's current state but also the C stack between the outermost runloop and the currently executing one. It would also have to recreate these call and runloop stacks when resuming the task. Otherwise the task would be ended as soon as the executing callback would be completed since instead of some C function called by an op in the bytecode, the scheduler is owner of the runloop.

@@

\section{Threads (5)}

@@

\begin{itemize}
\item replace thread\_data->interp\_lock usage by preallocated proxy objects
\item allow threads to create threads
\item allow multi level proxying (don't create a proxy for a proxy, unpack first)
\end{itemize}

>>>>

\section{Runloops}
\label{sec:Runloops}

@@

\chapter{Implementation of Hybrid Threads (13)}

The scheduler is the place where most of the green thread logic is implemented. It's functions use the prefix \textit{Parrot\_cx} and are located in \textit{src/scheduler.c}.

\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g. on interpreter startup and whenever a new thread is created. It creates the scheduler PMC and sets up alarm singal handling.

The scheduler is hooked into the system by replacing the call to execute the \textit{main\_sub} by a call to \textit{Parrot\_cx\_begin\_execution}. The latter creates the main task using \textit{main\_sub} as the task's code. This task is then put onto the run queue. The scheduler timer gets enabled and control given to \textit{Parrot\_cx\_outer\_runloop}.

\textit{Parrot\_cx\_outer\_runloop} is the core loop taking tasks from the run queue and executing them. Despite its name, it should not be confused with runloops discussed in section~\ref{sec:Runloops}. \textit{Parrot\_cx\_outer\_runloop} actually consists of two nested loops. The inner loop simply fetches tasks and executes as long as tasks are available. Even when all tasks are finished there may still be alarms pending which upon expiry would trigger new tasks to be scheduled. For this reason there's an outer loop checking for pending alarms and if there are putting the thread to sleep until the next alarm expires.

\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and executing it. Before calling the task, it checks if there are other tasks still in the queue. Only if other tasks are waiting, task preemption gets enabled. Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires. this is an important optimization reducing the runtime overhead of green threads in the important single tasking case essentially to 0.

To enable task preemption, \textit{Parrot\_cx\_enable\_preemption} sets a flag on the scheduler PMC and uses \textit{Parrot\_cx\_set\_scheduler\_alarm} to set an alarm at a point \textit{PARROT\_TASK\_SWITCH\_QUANTUM} milliseconds in the future.

When an alarm is triggered, Parrot only increments the global \textit{alarm\_serial}. This serial is checked for changes by \textit{Parrot\_alarm\_check} which is called by \textit{Parrot\_cx\_check\_scheduler}. This function in turn is called once for every \textit{branch} op. When the serial is different from the last check or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.

Since POSIX alarms can only be delivered to the main thread of a process, \textit{Parrot\_cx\_run\_scheduler} uses \textit{Parrot\_thread\_notify\_threads} to relay alarms to other thread's schedulers. For this purpose every thread has a pipe which \textit{Parrot\_thread\_notify\_threads} writes a dummy byte to. The next step for \textit{Parrot\_cx\_run\_scheduler} is to check for any expired alarms using \textit{Parrot\_cx\_check\_alarms}. An alarm contains a task as callback. On expiry this callback gets scheduled for immediate execution. As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it's the scheduler's preemption alarm. \textit{Parrot\_cx\_check\_quantum} is therefore used to check if the current task has used its assigned quantum and if it should be preempted.

Preemption of a task is implemented in \textit{Parrot\_cx\_preempt\_task}. It uses \textit{Parrot\_cx\_stop\_task} to create a continuation at the current point of execution and store it as the task's code. The task itself is then appended at the end of the run queue. The NULL opcode return value is propagated all the way up through the stack back to the \textit{branch} op checking for alarms. The op then recognizes this as the signal to stop processing and to exit the runloop.

The control flow ends up back at \textit{Parrot\_cx\_next\_task} at the point after executing the task at which it returns to \textit{Parrot\_cx\_outer\_runloop}.

The \textit{schedule} op code is used to schedule a new task from user code. It uses \textit{Parrot\_cx\_schedule\_task} which starts new worker threads if needed and possible and pushes the task on the target thread's scheduler's task queue. If the target thread previously was executing only a single task, its preemption has been disabled for optimization as described above. So in this case \textit{Parrot\_cx\_schedule\_task} has to enable preemption to give the new task a chance to run.

\textit{Parrot\_cx\_schedule\_immediate} is used in various places like at alarm expiry for putting a task to the head of the run queue and immediately causing preemption of the currently running task. It does this by setting \textit{SCHEDULER\_wake\_requested} and \textit{SCHEDULER\_resched\_requested} flags the same as when the preemption alarm expires. To be precise this mechanism leads to the current task being preempted at the next \textit{branch} op since this is the place when the mentioned flags will be checked.

Alarms can be registered using \textit{Parrot\_cx\_schedule\_alarm}. It puts the alarm in the appropriate place in the ordered alarm list and uses \textit{Parrot\_alarm\_set} which sets the actual alarm but only if there's not another alarm already set for an earlier time since there can only be one POSIX alarm pending at any time.

\textit{Parrot\_cx\_schedule\_sleep} is the actual implementation of the \textit{sleep} op. Like in\textit{Parrot\_cx\_preempt\_task}, \textit{Parrot\_cx\_stop\_task} is used to get an updated task at the current execution position, but instead of pushing this on the run queue, it is used as a callback for a newly set alarm. Again the NULL opcode is used to stop processing of the current task.

@@

Runloops are created whenever a part of Parrot's C internals starts to execute bytecode. This could be simply at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback. An exception handler could also be such a callback. Each runloop has a unique id. These ids are numbered from 0 and simply increased by 1 whenever a new runloop is created.

@@

To mitigate this, the \textit{reset\_runloop\_id\_counter(interp)} function is used by \textit{Parrot\_cx\_outer\_runloop} to reset the global runloop id counter back to 0 when resuming a task. This way the task retains the same runloop id over its life time.

If Parrot were to preempt a task while it is executing a nested runloop, it would have to somehow capture not only the interpreter's current state but also the C stack between the outermost runloop and the currently executing one. It would also have to recreate these call and runloop stacks when resuming the task. Otherwise the task would be ended as soon as the executing callback would be completed since instead of some C function called by an op in the bytecode, the scheduler is the owner of the runloop.

@@

\section{Threads (8)}

@@

\begin{itemize}
\item Windows implementation for timer/alarm handling
\item Replace thread\_data->interp\_lock usage by preallocated proxy objects
\item Allow threads to create threads
\item Allow multi level proxying (don't create a proxy for a proxy, unpack first)
\item Allow redistribution of tasks to other threads
\end{itemize}
